Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-09-15 paper.tex
Implementing latest comments + change to prop 2.2

Networks have become very powerful tools for representing and analysing
complex systems, with uses in a large array of fields. In network
science and statistics, they have been studies by various families of
models, from stochastic block models for detecting communities online
\citep{Latouche11}, to exponential random graph models (ERGMs) for
analysing the global trade network \citep{Setayesh22}, and mechanistic
models for investigating patterns in neural systems \citep{Betzel17}.

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that its degree distribution follows
a power law, that is the fraction of nodes with degree \(k\) is
proportional to \(k^{-\alpha}\), and therefore has a regularly varying
tail with tail index \(\alpha\). On the side against the claim is
\citet{Broido_2019who} compared the fits of a power law model against
that of several non-scale-free models to nearly a thousand networks,
only to find strong evidence for scale-freeness in four percent and weak
evidence in over half of the networks, thus claiming that scale-free
networks do not make up a majority in real networks. They compare the
fits of a power law model against that of several non-scale-free models
only finding strong evidence for scale-freeness in four percent and weak
evidence in fifty-two percent of the networks. On the other side of this
debate is \citet{Voitalov_2019} who disagrees and claims that these
networks are not nearly as rare and only appear to so as a result of an
unrealistic expectation of a power law without deviations or noise.
Additional evidence of these deviations from a power law is shown by
\citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is often
modelled well by a power law, while the tail is lighter than what is
implied by the body, albeit still regularly varying..

Nevertheless, most studies into the appropriateness of a power law for
the degrees of real networks, the aforementioned references included,
have been largely descriptive in the sense that no information about the
growth of the networks is revealed.

The popularity of using the power law for network degrees can be traced
back to the preferential attachment (PA) model popularised by
\citet{Barabasi99}. In the general model, as new nodes join the network,
an existing node with degree k gains edges at a rate proportional to
\(b(k)\), where b is a non-negative preference function.
\citet{Barabasi99} showed that, when \(b(k) = k + 1\), in the limit the
resulting degree distribution is regularly varying with index 3.
Subsequently, if a real network is shown to be scale-free, one can
loosely justify preferential attachment as the underlying mechanism of
its growth.

The model from \citet{Barabasi99} provided the foundations for various
generalisations.

\citet{krapivsky01} considered \(b(k) = (k+1)^\alpha\), and showed that
the degree distribution is not regularly varying (and therefore not
following the power law) when \(0<\alpha<1\), and when \(\alpha>1\) a
finite number of nodes end up with all edges after a certain point
resulting in a degenerate degree distribution. \citet{wang2022random}
returns to a linear preference function of the form
\(b(k) = k+\varepsilon\) but adds the possibility for reciprocal edges
to be sent, resulting in the joint distribution of in-degree and
out-degree being multivariate regularly varying and having the property
of hidden regular variation. \citet{rudas07} follows in the footsteps of
\citet{krapivsky01} , by considering a preferential attachment tree and
using theory from continuous branching processes, derives a limiting
degree distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

This paper aims to address the gap between the applied and theoretical
works, by asking if a network is assumed to come from a preferential
attachment model, can we use the degree distribution alone to directly
infer the parameters of the the preference function and learn about the
growth mechanisms? Moreover, proper consideration is given to the tail
of the degree distribution, because otherwise the effects of the largest
degrees, which correspond to the most influential nodes, deviating from
the power law will be discounted.

It is important to think about how we intend to consider the tail of the
degree distributions, because if not done correctly we may end up
essentially discounting the effects of the largest degrees (often the
most influential nodes) deviating from a power law.

For the above reason, we will use methods from discrete extreme value
theory, in particular those by \citet{shimura12}, who provided
theoretical guarantees for a discrete distribution to be regularly
varying or not. Using these results, we demonstrate how the tail of the
degree distribution is affected by \(b(\cdot)\), and subsequently
propose a class of preference functions that is tail realistic for real
networks. These analytical results enables the likelihood of the degree
distribution to be expressed in terms of the parameters of \(b(\cdot)\),
which in turn allows the underlying mechanism of the network, assumed to
grow according to preferential attachment, to be inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the preferential attachment model alongside the
theoretical results for the survival of the limiting degrees, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degrees
while remaining flexible up until a threshold. Section~\ref{sec-model}
utilises the proposed preference function and illustrates numerically
how the tail index of the degree distribution varies with the model
parameters. Section~\ref{sec-sim} consists of a simulation study where
networks are simulated from the proposed model for various parameter
combinations, demonstrating that the parameters can be recovered from
fitting the model to only the degree distribution.
Section~\ref{sec-real} fits the model to some real data and provides
posterior estimates for the preference function. Section~\ref{sec-conc}
concludes the article.

@@

The model that we will focus on in this paper is the General
Preferential Attachment model in \citet{rudas07} and is defined as
follows:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with tail index 2, and the Uniform Attachment (UA) model where
\(b(k)=c\) leading to a degree distribution that is not regularly
varying.

The survival function of the limiting degree distribution, called the
limiting survival hereafter, under condition \ref{eq-condb2} can be
analytically derived in the case where \(m=1\), this is presented below.

@@

\[
\Pr(\zeta(t+dt)=k+1|\zeta(t)=k) = b(k)dt + o(dt).
\] Now let \(\Upsilon(t)\) be the tree determined by \(\zeta(t)\) as
follows: \(\Upsilon(t)=\{\emptyset\}\) and \(\Upsilon(t)=G\) where each
existing node \(x\) in \(\Upsilon(t)\) gives birth to a child with rate
\(b(\mathrm{deg}(x, \Upsilon(t))\) independently of the other nodes
where \(\mathrm{deg}(x, \Upsilon(t))\) is the degree of node \(x\) in
the tree \(\Upsilon(t)\) at time \(t\).

@@

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
}\end{equation} where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\)
and \(\hat\rho\) is the Laplace transform of the density of the point
process associated with the pure birth process that corresponds to the
growth of an individual node, that is
\(\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t\)

@@

\begin{equation}\phantomsection\label{eq-surv-origin}{
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
}\end{equation}

@@

We are interested in how the tail behaviour of the discrete limiting
degree distribution is affected by the preference function \(b\).

For a distribution \(F\) with survival function \(\bar F\) and
\(k\in\mathbb Z^+\) let

@@

This allows us to show the following:

@@

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a regularly varying degree distribution, with tail index 2, by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be regularly varying we need
that the limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is
positive. The following proposition determines the class of functions
that will result in regular varying degree distributions.

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution is
regularly varying with tail index \(\lambda^{*}/c\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

From Proposition~\ref{prp-omega}, we have that: \[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = c>0.
\] Now, setting \(b_k = b(k)\) and \(a_k = k\):

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] Using the Stolz-Cesaro Theorem
\href{Appendix\%20-@sec-sup}{@thm-stolz},

\[
\lim_{k\rightarrow\infty}\frac{b_k}{a_k} = \lim_{k\rightarrow\infty}\frac{b(k)}{k} = c>0\qquad \square.
\]

@@

Using Proposition~\ref{prp-omega}, we can show that
\(\lim_{n\rightarrow\infty}\Omega(F,k)=\lambda^{*}/\beta\), meaning the
degree distribution obtained using this preference funnction is
regularly varying with tail index \(\lambda^*/\beta\).

@@

\section{A Preferential Attachment Model with Flexible Regular
Variation}\label{sec-model}

In the previous section, we found that using an asymptotically linear
preference function allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing regular
varation of the degrees. In this section, we demonstrate the flexibility
of the preference function in Equation~\ref{eq-pref}, with regard to the
tail behaviour of the limiting degree distribution. Using
Equation~\ref{eq-surv-origin}, the limiting survival is

@@

\caption{\label{fig-polylinsurv}The limiting survival, according to
various combinations of \((\alpha, \beta, \varepsilon)\) and \(k_0=20\)
of the proposed preferential attachment model.}

@@

The analytic form of the survival function in (\ref{eq-polysurv}),
offers a natural connection to the discrete version of the generalised
Pareto distribution (GP), providing a link to a well established
component of discrete extremes in the literature. Specifically,
connecting to the Integer GP (IGP) seen in \citet{Rohrbeck_2018} with
conditional survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)\) where \(\xi\) is the shape
parameter controlling the tail index.

@@

Therefore, \[{#eq-igp-est}
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution (for
large \(k_0^\alpha\)) is approximated by the
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is, the theoretical
conditional survival Equation~\ref{eq-polysurv} are shown in
Figure~\ref{fig-approx_surv} in colour and their IGP approximations
\textbf{?@eq-igp-est} are shown in grey. The approximation seems to hold
up fairly well even for large degrees and more so when \(\alpha\) is
larger.

@@

\caption{\label{fig-approx_surv}Theoretical conditional survivals (grey)
alongside their IGP approximations (coloured).}

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
regularly varying. Additionally the value of the shape parameter \(\xi\)
is shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model.

Through the connection to the IGP, fitting this model is almost
equivalent to fitting the IGP to the degrees and estimating its
parameters. However, instead of only describing the shape of the degree
distribution, we would also gain estimates for the shape of the
preference function, thus gaining a direct understanding the mechanisms
underlying the growth of the network.

@@

where \(\text{B}(\cdot,\cdot)\) is the the beta function, and \(l\ge0\)
is a quantity that allows truncating the data such that the minimum
degree is \(l\). This will allow the model to be fitted whilst ignoring
the influence of the lower degrees (those less than \(l\)) as the model
does not capture the behaviour well at the lower degrees since
\citet{rudas07} only provides results for the case of a preferential
attachment tree.

@@

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts and using the likelihood from the previous section alongisde the
priors:

@@

where Gamma(\(a\), \(b\)) is the Gamma distribution with shape \(a\) and
rate \(b\), and U(\(a\), \(b\)) is uniform distribution with lower and
upper bounds a and b, to obtain a posterior distribution, up to the
proportionality constant. Posterior samples can then be obtained by an
adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm.
For these simulated networks \(l=0\).

@@

\caption{\label{fig-rec1}Posterior estimates of survival function for
data simulated from the proposed model with various combinations of
\((\alpha,\beta,\varepsilon)\) and \(k_0=20\).}

@@

\caption{\label{fig-rec2}: Posterior estimates of paramters for data
simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

@@

In this section, we fit the proposed model to the degree distributions
of various real networks and learn about the mechanics of their growth.
While we also compare the fit to that of the mixture distribution by
\citet{Lee24} we note that the proposed method has the additional
benefit of learning directly about the growth of a network from the
inference results. The data consists of 12 networks sourced from
\href{konect.cc}{KONECT} and the
\href{https://networkrepository.com}{Network Data Repository}\citep{nr}:

@@

\caption{\label{fig-real1}Posterior estimates (solid red) of survival
for several real data sets and their 95\% credible intervals (dotted
red).}

@@

Figure~\ref{fig-real1} displays the posterior estimates of the survival
function for various data sets, obtained from fitting the GPA model and
the Zipf-IGP mixture model from \citet{Lee24}. In most cases, the GPA
model gives a similar fit to the Zipf-IGP model but where the GPA model
fits well we gain additional information about the preference function
assuming that the network evolved according the the GPA scheme.

Figure~\ref{fig-shapes} shows the posterior of the shape parameter
\(\xi\) obtained from the Zipf-IGP model alongside the posterior of the
equivalent shape parameter \(\beta/\lambda^*\) obtained from fitting the
GPA model. Generally, the GPA model performs similarly to the Zipf-IGP
when estimating the tail behaviour of the degree distribution. In the
cases of substantial discrepancies, it is either because the GPA model
fits the tail better than the Zipf-IGP model does, or because of the
threshold being estimated as too low forcing almost all of the data to
be modelled by the linear part of the GPA. This again shows the effects
that small degrees have on this model.

@@

\caption{\label{fig-shapes}Posterior estimates (solid red) of survival
for several real data sets and their 95\% credible intervals (dotted
red).}

@@

\caption{\label{fig-pa}Posterior estimate for preference function
(solid) with 95\% credible interval (dashed) on log-log scale.}

@@

One limitation of this method is that the lowest degrees needed to be
truncated as they had a very large effect on the fit of the model as a
result of using theory developed for trees and applying it to general
networks. Future work could apply theory developed for general networks
using a similar method to this, allowing us to compare the results here
something that is more accurate. This could include fixing the
out-degree of new nodes at a constant greater than one, or allowing the
out-degree of new nodes to vary.

>>>>

Networks have become powerful tools for representing and analysing
complex systems, with uses in a large array of fields. In network
science and statistics, they have been studied by various families of
models, from stochastic block models for detecting communities online
\citep{Latouche11}, to exponential random graph models (ERGMs) for
analysing the global trade network \citep{Setayesh22}, and mechanistic
models for investigating patterns in neural systems \citep{Betzel17}.

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that its degree distribution follows
a power law, that is the fraction of nodes with degree \(k\) is
proportional to \(k^{-\alpha}\), and therefore has a regularly varying
tail with tail index \(\alpha\). On the side against the claim is
\citet{Broido_2019} who compared the fits of a power law model against
that of several non-scale-free models to nearly a thousand networks,
only to find strong evidence for scale-freeness in four percent and weak
evidence in over half of the networks, thus claiming that scale-free
networks do not make up a majority in real networks. On the other side
of this debate is \citet{Voitalov_2019} who disagrees and claims that
these networks are not nearly as rare and only appear to so as a result
of an unrealistic expectation of a power law without deviations or
noise. Additional evidence of these deviations from a power law is shown
by \citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is often
modelled well by a power law, while the tail is lighter than what is
implied by the body, albeit still regularly varying. Nevertheless, most
studies into the appropriateness of a power law for the degrees of real
networks, the aforementioned references included, have been largely
descriptive in the sense that no information about the growth of the
networks is revealed.

The popularity of using the power law for network degrees can be traced
back to the preferential attachment (PA) model popularised by
\citet{Barabasi99}. In the general model, as new nodes join the network,
an existing node with degree \(k\) gains edges at a rate proportional to
\(b(k)\), where \(b(\cdot)\) is a non-negative preference function.
\citet{Barabasi99} showed that, when \(b(k) = k + 1\), in the limit the
resulting degree distribution is regularly varying with index 2.
Subsequently, if a real network is shown to be scale-free, one can
loosely justify PA as the underlying mechanism of its growth.

The model from \citet{Barabasi99} provided the foundations for various
generalisations --- \citet{krapivsky01} considered
\(b(k) = (k+1)^\alpha\), and showed that the degree distribution is not
regularly varying (and therefore not following the power law) when
\(0<\alpha<1\), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, resulting in the joint distribution of
in-degree and out-degree being multivariate regularly varying and having
the property of hidden regular variation. \citet{rudas07} follows in the
footsteps of \citet{krapivsky01}, by considering a PA tree and using
theory from continuous branching processes, derives a limiting degree
distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

This paper aims to address the gap between the applied and theoretical
works, by asking if a network is assumed to come from a PA model, can we
use the degree distribution alone to directly infer the parameters of
the the preference function and learn about the growth mechanisms?
Moreover, proper consideration is given to the tail of the degree
distribution, because otherwise the effects of the largest degrees,
which correspond to the most influential nodes, deviating from the power
law will be discounted.

As the degrees of networks are discrete values, we will use methods from
discrete extreme value theory, in particular those by \citet{shimura12},
who provided theoretical guarantees for a discrete distribution to be
regularly varying or not. Using these results, we demonstrate how the
tail of the degree distribution is affected by \(b(\cdot)\), and
subsequently propose a class of preference functions that is tail
realistic for real networks. These analytical results enables the
likelihood of the degree distribution to be expressed in terms of the
parameters of \(b(\cdot)\), which in turn allows the underlying
mechanism of the network, assumed to grow according to PA, to be
inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the PA model alongside the theoretical results
for the survival function of the limiting degree distribution, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degree
distribution while remaining flexible up until a threshold.
Section~\ref{sec-model} utilises the proposed preference function and
illustrates numerically how the tail index of the degree distribution
varies with the model parameters. The simulation study in
Section~\ref{sec-sim} demonstrates that the parameters can be recovered
from fitting the model to only the degree distribution.
Section~\ref{sec-real} fits the model to some real data and provides
posterior estimates for the preference function. Section~\ref{sec-conc}
provides a discussion of this paper and possible avenues for future
work.

@@

The model that we will focus on in this paper is the General
Preferential Attachment (GPA) model in \citet{rudas07} and is defined as
follows:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+1\), which in the limit of \(t\rightarrow \infty\) leads to a
power-law degree distribution with tail index 2, and the Uniform
Attachment (UA) model where \(b(k)=c\) leading to a degree distribution
that is not regularly varying.

The survival function of the limiting degree distribution, called the
limiting survival hereafter, under condition \ref{eq-condb2} can be
analytically derived in the case where \(m=1\), which is presented
below.

@@

\[
\Pr(\zeta(t+\text{d}t)=k+1|\zeta(t)=k) = b(k)\text{d}t + o(\text{d}t).
\] Now let \(\Upsilon(t)\) be the tree determined by \(\zeta(t)\) as
follows: \(\Upsilon(t)=\{\emptyset\}\) and \(\Upsilon(t)=G\) where each
existing node \(x\) in \(\Upsilon(t)\) gives birth to a child with rate
\(b(\mathrm{deg}(x, \Upsilon(t))\) independently of the other nodes
where \(\mathrm{deg}(x, \Upsilon(t))\) is the degree of node \(x\) in
the tree \(\Upsilon(t)\) at time \(t\), denote by
\(\Upsilon(t)_{\downarrow x}\) the tree when treating node \(x\) as the
root.

@@

\begin{equation}\phantomsection\label{eq-survlim}{
\lim*{t*\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum{x\in\Upsilon(t)}\varphi(\Upsilon(t)\_{\downarrow x}) = \lambda\^\* \int\_0\infty e{-\lambda\^\* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t 
}\end{equation}

where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\) and \(\hat\rho\)
is the Laplace transform of the density of the point process associated
with the pure birth process that corresponds to the growth of an
individual node, that is
\(\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t\).

@@

\begin{equation}\phantomsection\label{eq-surv-origin}{
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]\text{d}t}{\int_0^\infty e^{-\lambda^* t}\text{d}t} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
}\end{equation}

@@

We are interested in how the tail behaviour of the discrete limiting
degree distribution is affected by the preference function \(b\). In
order to study this, we use the quantity \(\Omega(F, k)\) as stated in
\citet{shimura12} and is repeated below:

For a discrete distribution \(F\) with survival function \(\bar F\) and
\(k\in\mathbb Z^+\) let

@@

The following proposition gives the limiting behaviour of
\(\Omega(F,k)\) when \$F\$ is a limiting degree distribution resulting
from the GPA model with preference function \(b(\cdot)\).

@@

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition~\ref{prp-omega} also aligns with the fact
that BA model produces a regularly varying degree distribution, with
tail index 2, by considering the preference function
\(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\), leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be regularly varying we need
that the limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is
positive. The following proposition determines the class of functions
that will result in regular varying degree distributions.

@@

The limiting survival of a GPA model is regularly varying if and only if
the preference function is \emph{eventually} of the form:

\begin{align*}
b(k) &= \beta k + r(k),\\
&\text{and}\\
r(k+1)-r(k)&\rightarrow 0,
\end{align*}

and if the preference function is of this form then the tail index is
exactly \(\lambda^*/\beta\) where \(\hat\rho(\lambda\^\*) = 1\).

@@

Letting \(b(k) = \beta k  + o(k)\) and using Proposition~\ref{prp-omega}
we have that:

\[
\lim_{k\rightarrow\infty}\Omega(F,k) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} = \lim_{k\rightarrow\infty} \frac{\beta + o(k+1) - o(k)}{\lambda^*} = \frac{\beta}{\lambda^*},
\]

and thus the tail index of the limiting survival must be
\(\lambda^*/\beta\).

\hfill\break
If we now assume that the limiting survival is regularly varying with
tail index \(\delta\), we must have that:

\[
\lim_{k\rightarrow\infty}\Omega(F,k) = \delta
\]

and again by Proposition~\ref{prp-omega}

\[
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} = \delta
\]

Letting \(b_k=b(k)\) and \(a_k = k\) we can use the Stolz-Cesàro theorem
(Theorem~\ref{thm-stolz}) to obtain:

\[
\lim_{k\rightarrow\infty}\frac{b_{k+1}-b_k}{a_{k+1}-a_k} = \delta \implies \lim_{k\rightarrow\infty}\frac{b_k}{a_k} = \delta
\]

and so

\[
\lim_{k\rightarrow\infty}\frac{b(k)}{k} = \delta
\]

which \(b(k) = \beta k + o(k)\) fulfills.\(\square\)

@@

Using Proposition~\ref{prp-omega}, we can show that
\(\lim_{n\rightarrow\infty}\Omega(F,k)=\lambda^{*}/\beta\), meaning the
degree distribution obtained using this preference function is regularly
varying with tail index \(\lambda^*/\beta\).

@@

\section{A PA Model with Flexible Regular Variation}\label{sec-model}

In the previous section, we found that using an asymptotically linear
preference function allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing regular
variation of the degrees. In this section, we demonstrate the
flexibility of the preference function in Equation~\ref{eq-pref}, with
regard to the tail behaviour of the limiting degree distribution. Using
Equation~\ref{eq-surv-origin}, the limiting survival is

@@

\caption{\label{fig-polylinsurv}The limiting survival, according to
various combinations of \((\alpha, \beta, \varepsilon)\) and \(k_0=20\)
of the proposed PA model.}

@@

The analytic form of the survival function in (\ref{eq-polysurv}),
offers a natural connection to the discrete version of the generalised
Pareto (GP) distribution , providing a link to a well established
component of discrete extremes in the literature. Specifically, it is
connected to the Integer GP (IGP) distribution seen in
\citet{Rohrbeck_2018} with conditional survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)\) where \(\xi\) is the shape
parameter and reciprocal of the tail index.

@@

Therefore, \begin{equation}\phantomsection\label{eq-igp-est}{
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
}\end{equation} meaning that for \(k\ge k_0\) the limiting degree
distribution (for large \(k_0^\alpha\)) is approximated by the
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\)
distribution.

To assess how close of an approximation this is, the theoretical
conditional survival Equation~\ref{eq-polysurv} are shown in
Figure~\ref{fig-approx_surv} in colour and their IGP approximations
Equation~\ref{eq-igp-est} are shown in grey. The approximation holds up
fairly well even for large degrees and more so when \(\alpha\) is
larger.

@@

\caption{\label{fig-approx_surv}Theoretical conditional survivals (grey)
of the proposed model alongside their IGP approximations (coloured).}

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP distribution is positive and thus the
distribution is regularly varying. Additionally the shape parameter
\(\xi\) is shown in Figure~\ref{fig-polyheat} for various parameter
choices. The darker and lighter regions on the heat maps correspond to a
heavier and a lighter tail, respectively, and the red dashed line shows
combinations of \(\alpha\) and \(\beta\) that produce a limiting degree
distribution with the same tail index as the BA model.

Through the connection implied by Equation~\ref{eq-igp-est}, fitting the
proposed model is almost equivalent to fitting the IGP distribution to
the degrees and estimating its parameters. However, instead of only
describing the shape of the degree distribution, we would also gain
estimates for the shape of the preference function, thus gaining a
direct understanding the mechanisms underlying the growth of the
network.

@@

where \(\text{B}(\cdot,\cdot)\) is the the beta function,
\(\pmb \theta = (\alpha, \varepsilon, k_0,\beta)\), and \(l\ge0\) is a
quantity that allows truncating the data such that the minimum degree is
\(l\). This will allow the model to be fitted whilst ignoring the
influence of the lower degrees (those less than \(l\)) as the model does
not capture the behaviour well at the lower degrees, since
\citet{rudas07} only provides results for the case of a preferential
attachment tree.

@@

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts and using the likelihood from the previous section alongside the
priors:

@@

where Gamma(\(a\), \(b\)) is the gamma distribution with shape \(a\) and
rate \(b\), and U(\(a\), \(b\)) is uniform distribution with lower and
upper bounds \(a\) and \(b\), to obtain a posterior distribution, up to
the proportionality constant. Posterior samples can then be obtained by
an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC)
algorithm. For these simulated networks \(l=0\).

@@

\caption{\label{fig-rec1}Empirical (black dots) and fitted (red line)
survival functions for data simulated from the proposed model with
various combinations of \((\alpha,\beta,\epsilon)\) and \(k_0=20\). The
95\% credible (dashed lines) are included but too narrow to be seen
clearly.}

@@

\caption{\label{fig-rec2}: Posterior densities of parameters for data
simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\). True parameter
values shown with black dots.}

@@

In this subsection, we fit the proposed model to the degree
distributions of various real networks and learn about the mechanics of
their growth. While we also compare the fit to that of the mixture
distribution by \citet{Lee24} we note that the proposed method has the
additional benefit of learning directly about the growth of a network
from the inference results. The data consists of 12 networks sourced
from \href{konect.cc}{KONECT} and the
\href{https://networkrepository.com}{Network Data Repository}\citep{nr}:

@@

\caption{\label{fig-real1}Empirical (black dots) and posterior medians
(solid red) of the fitted survival function for several real data sets
and their 95\% credible intervals (dotted red).}

@@

Figure~\ref{fig-real1} displays the posterior estimates of the survival
function for various data sets, obtained from fitting the GPA model and
the Zipf-IGP mixture model from \citet{Lee24}. In most cases, the GPA
model gives a similar fit to the Zipf-IGP model but where the GPA model
fits well we gain additional information about the preference function,
assuming that the network evolved according the the GPA scheme.

\caption{Posterior densities of the parameters of the proposed model
fitted to real data.}

Figure~\ref{fig-shapes} shows the posterior of the shape parameter
\(\xi\) obtained from the Zipf-IGP model alongside the posterior of the
equivalent shape parameter \(\beta/\lambda^*\) obtained from fitting the
GPA model. Generally, the GPA model performs similarly to the Zipf-IGP
when estimating the tail behaviour of the degree distribution. In the
cases of substantial discrepancies, it is either because the GPA model
fits the tail better than the Zipf-IGP model does, or because of the
threshold being too low, forcing almost all of the data to be modelled
by the linear part of the GPA. This again shows the effects that small
degrees have on this model.

@@

\caption{\label{fig-shapes}Posterior estimates of shape parameter of
Zipf-IGP distribution (right) and the analagous quantity obtained from
fitting the proposed model (left).}

@@

\caption{\label{fig-pa}Posterior median for the preference function
(solid) with 95\% credible interval (dashed) on log-log scale.}

@@

This paper contributes to the understanding of the relationship between
the mechanism underlying a networks growth and the resulting degree
distribution. As well as, demonstrating that under certain conditions
information about this mechanism can be garnered from the degrees alone.
Perhaps in future, something similar could be done by looking at another
statistic for the network (e.g.~the triangle distribution) and using it
in combination with the degrees to gain further insights into the growth
mechanism.

One limitation of this method is that the lowest degrees needed to be
truncated as they had a very large effect on the fit of the model, as a
result of using theory developed for trees and applying it to general
networks. Future work could apply theory developed for general networks
using a similar method to this, allowing us to compare the results here
something that is more accurate. This could include fixing the
out-degree of new nodes at a constant greater than one, or allowing the
out-degree of new nodes to vary.
